### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
removing residual norm figure from spn estimator comparison

% on the SPn equations which I am effectively following here because I
% was feeling lazy. I have verified its correctness.

@@

%%---------------------------------------------------------------------------%%
\section{Basic Preconditioning for the $SP_N$ Equations}
\label{sec:spn_spectral_analysis}
Previous work in applying MCSA to the thermal radiation diffusion
equations had success in using Jacobi preconditioning
\cite{evans_monte_2012}. In this section, we will compute the spectral
radius of the iteration matrix formed by the $SP_N$ equations while
parametrically varying the $SP_N$ order, the $P_N$ order, the number
of energy groups, and upscattering/downscattering between energy
groups for Jacobi preconditioned problems to ensure that this type of
preconditioning will be applicable. For each problem in the parameter
study, the values given in Table~\ref{tab:spn_fixed_parameters} were
fixed for each spatial location and energy group.

@@

\subsection{ILUT Preconditioning}
\label{subsec:spn_ilut_preconditioning}
ILUT preconditioning is chosen due to its applicability to problems
that have a dominating elliptic compoenent such as the fuel assembly
problem and because it is easily formulated algebraicly. Incomplete
lower-upper (ILU) factorizations of the linear operator can be used a
simple mechanism to form an approximate inverse of a
preconditioner. To build the factorization, the sparse upper and lower
triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are computed such
that the residual matrix formed by the factorization
\cite{saad_iterative_2003}:
\begin{equation}
  \mathbf{R} = \mathbf{L} \mathbf{U} - \mathbf{A} \:,
  \label{eq:ilu_residual_matrix}
\end{equation}
has a specified sparsity pattern and element magnitude. When a
magnitude threshold is used, elements generated in the factors below
that magnitude are dropped, resulting in ILU threshold (ILUT)
preconditioning. The sparsity pattern in this case is determined from
the input matrix to be preconditioned and the number of elements
maintained in the factor is specified by a fill level parameter. A
fill level of 1 will generate the same number of elements as the
sparsity pattern of the input matrix while a fill level of 2 will
contain twice as many elements in the factorization, resulting in a
better representation of the true LU factorization. The inverse of the
lower and upper triangular factors may then be easily inverted by
means of simple elimination to produce the preconditioner. For MCSA,
$\mathbf{L}^{-1}$ will be used on the left and $\mathbf{U}^{-1}$ on
the right to precondition the system.

@@

\subsection{Sparse Approximate Inverse Preconditioning}
\label{subsec:spn_spainv_preconditioning}
SPAINV is chosen because it the true inverse of the preconditioner is
byproduct of the algorithm and the algorithm seeks to maintain the
sparsity of the inverse preconditioner, potentially alleviating some
of the issues observed with ILUT. Explicit formation of the
preconditioner inverse is a requirement of the current MCSA
preconditioning strategy. As a result, although ILUT preconditioning
provided excellent iterative performance, a significant time penalty
is paid to extract the inverse of the triangular factors through $N$
triangular solves. In addition, this explicit inversion yielded a
composite linear operator that was dense, requiring modification of
the system in order to achieve better scalability and Monte Carlo
performance. Sparse approximate inverse (SPAINV) preconditioning is a
technique that may potentially alleviate both of these constraints at
the cost of reduced iterative performance. The method produces a
sparse approximation of the inverse matrix directly by minimizing the
Frobenius norm of the residual matrix \cite{saad_iterative_2003}:
\begin{equation}
  || \mathbf{I} - \mathbf{A} \mathbf{M} ||^2_F =
  \sum_{j=1}^N ||\mathbf{e}_j - \mathbf{A} \mathbf{m}_j||^2_2 \:,
  \label{eq:frobenius_norm_min}
\end{equation}
where the inverse preconditioning matrix $\mathbf{M}$ minimizes the
norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
$j^{th}$ columns of the identity and preconditioning matrices
respectively. Here, $\mathbf{M}$ is the actual inverse of the
preconditioner and is formed directly by the algorithm. On the right
hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
represented as a effective sum of $N$ linear system residual norms.

@@

\subsection{Applying the Reduced Domain Approximation}
\label{subsec:spn_prec_rda}
For each preconditioning technique presented convergence was achieved
for the single fuel assembly problem. However, a primary concern is
the number of non-zero states in each row of the system generated by
the explicit preconditioning strategy. In many cases, orders of
magnitude more matrix elements were generated resulting in poor
scalability for domain decomposed algorithms and overall performance
issues for Monte Carlo. As outlined in
\S~\ref{subsec:reduced_domain_approximation}, the reduced domain
approximation may be used as a mechanism to potentially alleviate this
problem by filtering elements of the composite matrix in each row that
fall below a certain threshold value or by maintaining the largest $N$
elements in each row where $N$ is a designated fill level.

For the ILUT, SPAINV, and SPAINV-ILUT preconditioning strategies the
reduced domain approximation will be applied to reduce the density of
the composite linear operator to more manageable levels. For each
preconditioner, the parameters that achieved the best quality metric
results from the previous analysis were used. These correspond to ILUT
parameters of a fill level of 5.0 and a drop tolerance of \sn{1}{-5},
SPAINV parameters of a sparsity level of 4 and a threshold of 0.1 and
SPAINV-ILUT with ILUT parameters of a fill level of 4.0 and a drop
tolerance of \sn{1}{-5} and SPAINV parameters of a sparsity level of 1
and a threshold of 1.0. The reduced domain threshold was set to
\sn{1}{-10} in order to eliminate any exceedingly small values from
the matrix (often this is simply removing non-zero elements within the
floating point tolerance of zero). The reduced domain fill level was
then varied, starting with the largest non-zero entries per row value
observed for each of the preconditioning types in order to assess its
effects relative to the case where no reduced domain approximation was
applied.

Figure~\ref{fig:rda_iterations} gives the number of iterations
required to converge for each preconditioner type as a function of
reduced domain fill level. Figure~\ref{fig:rda_quality} gives the
corresponding quality metric for each data point where the number of
non-zero entries used to compute the metric is equivalent to the
reduced domain fill level. We first note that SPAINV preconditioning
alone is significantly more sensitive to the reduction in domain size
over the ILUT-based methods, although the preconditioner was of
$O(100)$ non-zero entries per row without any approximation
applied. For the ILUT-based methods, performance was significantly
better with convergence achieved in less than 40 MCSA iterations with
only 10 non-zero entries in each row (vs. 7 non-zero entries for the
case with no preconditioning). SPAINV-ILUT iterative performance was
marginally better than ILUT alone for all reduced domain fill
levels. However, it should be noted that the ILUT level of fill used
for the ILUT only calculations was set to 4 for this case while the
SPAINV-ILUT preconditioner used an ILUT level of fill of 5 and
therefore the marginally better performance is more likely a result of
this addition of fill rather than the extra SPAINV
preconditioning. Looking at the quality metric data, we see a
reduction in the quality metric as a function of the reduced domain
fill level, achieving 2 orders of magnitude reduction in the quality
metric for the ILUT-based preconditioning methods.

@@

Figure~\ref{fig:relax_iters} gives the number of iterations required
to converge the fuel assembly problem for a 1-group SP1 discretization
using varying combinations of the relaxation parameters. First, both
parameters were fixed at the base case of 1 while the other parameter
was varied as shown in the left-hand side plots of
Figure~\ref{fig:relax_iters}. For the Richardson relaxation parameter,
using a value larger than 1 gave better iterative performance up to a
point, effectively providing a stronger extrapolation using the
residual at each iteration. For the Neumann relaxation parameter, a
value of less than 1 is observed to be ideal. Although initially
counter-intuitive, the fact that the correction computed by the Monte
Carlo solver has a stochastic error associated with it means that by
using a Neumann relaxation parameter less than 1, the correction and
its error are effectively dampened to improve iterative
performance. Considering the CPU time required to converge a single
eigenvalue iteration, similar results are also observed for the
relaxation parameters as given by Figure~\ref{fig:relax_time}. For the
base cases given by the plots on the left, it was found the a
Richardson relaxation parameter of 1.1 and a Neumann relaxation
parameter of 0.7 provided the fastest CPU time for convergence. Fixing
each parameter at these new values, the calculations were repeated as
shown for the plots on the right hand sides of both
Figures~\ref{fig:relax_iters} and \ref{fig:relax_time}. For each
repeated timing calculation, it was found that the same combination of
relaxation parameters found in the base cases performed the best
although they did not necessarily have the best iterative performance.

@@

\subsection{MCSA Estimator and Fixed Point Iteration Analysis}
\label{subsec:spn_estimator_comparison}
With convergence obtained for the fuel assembly problem, the iterative
performance of MCSA will next be studied with both the collision and
expected value estimators using the best preconditioning and
relaxation parameter combinations found by the previous
analysis. Additionally, both the Richardson-based MCSA iteration given
by Eq~(\ref{eq:mcsa}) and the minimal residual-based MCSA iteration
given by Eq~(\ref{eq:mcsa_min_res}) will be used. Although the
expected value estimator achieved the best iterative performance in
the previous comparison and analysis for the simple Poisson problem in
\S~\ref{sec:parameter_estimator_analysis}, it is possible that the
collision estimator may perform better from a timing perspective due
to the sparsity lost by explicit preconditioning. The deterministic
component of the expected value estimator that couples the current
state to other local states through the iteration matrix stencil
requires application of the estimator to potentially several orders of
magnitude more states at every transition event.

@@

Figure~\ref{fig:spn_estimator_iters} gives the iteration count results
and Figure~\ref{fig:spn_estimator_time} gives the timing
results. Remarkably, although the fuel assembly problem is
significantly more complicated than the simple Poisson problem studied
in \S~\ref{sec:parameter_estimator_analysis}, the results here are
largely the same with Figure~\ref{fig:spn_estimator_iters} and
Figure~\ref{fig:estimator_nh_iters} having identical qualitative
behavior. Using the expected value estimator with both fixed point
iterations again creates a relative insensitivity to the number of
histories used to compute the correction. Marginally better iterative
and timing performance was observed for the expected value estimator
when using the Richardson iteration. Better timing performance is
expected as computing the extrapolation parameter in the minimal
residual iteration requires several additional parallel
operations. Compared to the collision estimator, the minimal residual
iteration permits fewer stochastic histories to be used with each MCSA
iteration while still maintaining good iterative performance. With
respect to CPU time, there is a more pronounced effect due to the
larger density of states created by the explicit ILUT
preconditioning. Using a reduced domain fill level of 100 creates a
situation where the expected value estimator is significantly more
expensive per history to compute that the collision estimator due to
the coupling of states.

@@

From both Figure~\ref{fig:spn_estimator_iters} and
Figure~\ref{fig:spn_estimator_time} we see that both estimators and
fixed point iterations have comparable iterative and timing
performance when 1,000 histories are used at each MCSA
iteration. Using 1,000 histories per MCSA iteration, the infinity norm
of the residual vector is presented at every MCSA iteration for both
estimators and both fixed point iterations in
Figure~\ref{fig:spn_estimator_convergence} as an additional means of
comparison. At this number of histories, all MCSA combinations
converge the fuel assembly problem monotonically at approximately the
same rate, giving little reason to choose one over the other. The end
result is that both estimators provide a viable method for getting
good MCSA iterative performance with comparable timing
performance. For the collision estimator, using around 20,000
stochastic histories at every iteration gave the best iterative
performance while 500 histories per iteration gave the best iterative
performance when using the expected value estimator. For purely serial
computational performance, there is not a distinguishing feature for
the fuel assembly problem presented that would cause one to choose one
of the estimators and fixed point iterations over the other as they
all have similar minimum CPU times over the range of values tested.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/estimator_convergence.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm as a function of MCSA
      iteration with 1,000 stochastic histories per iteration.}
    \textit{Both the collision and expected value estimators were used
      with the adjoint Monte Carlo solver to compute the correction at
      each iteration. Each estimator was used with the Richardson and
      minimal residual fixed point iterations within an MCSA
      iteration.}}
  \label{fig:spn_estimator_convergence}
\end{figure}

@@

%%---------------------------------------------------------------------------%%
\section{MCSA Verification\ }
\label{sec:spn_mcsa_verification}
Through the numerical studies presented in this chapter, we have
developed a set of preconditioning techniques that permit MCSA to be
used with the $SP_N$ form of the neutron transport equation and
applied to a difficult fuel assembly criticality problem. Along with
these preconditioning techniques, a set a varying MCSA parameters were
analyzed to find those that yielded the best performance for this
particular problem. In addition, several steps were taken to mitigate
the dense composite linear operators that arise when performing
explicit MCSA preconditioning necessary for achieve convergence. Using
these results, we know compare MCSA directly to conventional linear
solvers that would be typically used to solve the $SP_N$ form of the
transport equation as presented here. In particular, we will use a
pair of Krylov solvers with which we will compare numerical results to
verify MCSA solutions the implementation used for this work.

@@

Table~\ref{tab:serial_differences_g1},
Table~\ref{tab:serial_differences_g2}, and
Table~\ref{tab:serial_differences_g4} give the infinity norms and
2-norms of the flux relative difference vectors given by
Eq~(\ref{eq:relative_difference_vector}) for all calculations and all
energy groups. In general, we see good agreement in the two norm
between the MCSA variants and the GMRES calculation, meaning that they
have approximately the same variation from the BiCGStab
calculation. We do note larger infinity norm values of this vector for
some of the MCSA variants (i.e. group 1 in the 2-group calculation),
however this large maximum value was not a global phenomenon with good
agreement in the 2-norm for the calculations. Based on the good
agreement between MCSA and GMRES for the relative difference of the
group fluxes and the eigenvalue results presented here, the MCSA
implementation should be deemed correct for the fuel assembly problem
for serial calculations with multiple energy groups.

@@

%%---------------------------------------------------------------------------%%
\section{MCSA Performance Comparison to Conventional Methods\ }
\label{sec:spn_comparison}
In the previous section, numerical results using MCSA for the fuel
assembly calculation were directly compared with production Krylov
methods in order to verify their correctness. Deemed correct,
performance of the algorithm for serial computations will now be
addressed. With the same set of solver parameters used for
verification caluclation, MCSA will now be compared to the same Krylov
solvers in terms of both iterative performance and CPU timing. A
Richardson iteration solution will also be used in order to assess the
acceleration provided by the residual Monte Carlo component of MCSA.

To compare iterative performance, the number of linear solver
iterations required to converge each eigenvalue iteration was recorded
and then averaged over all eigenvalue iterations for each
solver. Table~\ref{tab:spn_comparison_iterations} gives the average
number of linear solver iterations required to converge a single
eigenvalue iteration of the fuel assembly problem as a function of the
number of energy groups. In general, the iterative performance of all
the solvers tested aside from the Richardson iteration is comparable
with BiCGStab performing the best. We expect this because not only are
the multigroup $SP_N$ equations positive-definite, but they are also
block-symmetric and therefore we expect the conjugate gradient-based
methods to perform well due to the structure of the projection
scheme. It important to note that the iterative performance of each
solver is not a strong function of the number of energy groups in the
problem (and therefore the problem size). Even more important,
although BiCGStab produced the best iterative results, MCSA did
perform better than GMRES in many cases with the Richardson iteration
clearly accelerated by the Monte Carlo solve. When compared to the
preconditioned Richardson iteration, the acceleration provided by MCSA
is clear with 3-4 times fewer iterations required for convergence.

@@

\begin{itemize}
\item MCSA can solve the asymmetric system generated by the $SP_N$
  equations
\item MCSA has been incorporated into the Exnihilo neutronics
  production code base developed at Oak Ridge National Laboratory
\item Simple Jacobi-based preconditioning reduces the spectral radius
  of the $SP_N$ system below unity, however, this alone is not
  sufficient for MCSA convergence
\item As the spectral radius approaches unity, MCSA was demonstrated
  to break down with more stochastic histories required for
  convergence and more CPU time required per history
\item Light water reactor problems are difficult to solve with MCSA as
  they have large spectral radii due to the neutron scattering in the
  moderator
\item Advanced algebraic preconditioning strategies were applied to
  the $SP_N$ equations to obtain convergence with ILUT chosen for
  subsequent investigations
\item The reduced domain approximation was required to mitigate the
  memory and performance constraints of the explicit preconditioning
  strategy although both remain significant obstacles
\item MCSA relaxation parameters were developed and observed to
  enhance the time to solution for the fuel assembly criticality
  problem
\item MCSA was verified to produce the same flux distribution and
  k-eigenvalue for the fuel assembly as production Krylov methods
\item MCSA was observed to converge in fewer iterations per eigenvalue
  iteration than GMRES for the fuel assembly criticality problem and
  more than Bi-CGStab using the same preconditioning
\item The explicit preconditioning strategy required to overcome the
  MCSA spectral radius restriction creates runtimes $O(100)$ larger
  than the production Krylov methods for the fuel assembly criticality
  problem
\end{itemize}

>>>>

% on the SPn equations.

@@

%%---------------------------------------------------------------------------%%
\section{Basic Preconditioning for the $SP_N$ Equations}
\label{sec:spn_spectral_analysis}
Previous work in applying MCSA to the thermal radiation diffusion
equations had success in using Jacobi preconditioning
\cite{evans_monte_2012}. In this section, we will compute the spectral
radius of the iteration matrix formed by the $SP_N$ equations while
parametrically varying the $SP_N$ order, the $P_N$ order, the number
of energy groups, and upscattering/downscattering between energy
groups for Jacobi preconditioned problems to ensure that this type of
preconditioning will be applicable to the fuel assembly problem. For
each problem in the parameter study, the values given in
Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
location and energy group.

@@

\subsection{ILUT Preconditioning}
\label{subsec:spn_ilut_preconditioning}
ILUT preconditioning is chosen first due to its applicability to
problems that have a dominating elliptic component such as the fuel
assembly problem or other problems with a high scattering to
absorption ratio. Incomplete lower-upper (ILU) factorizations of the
linear operator can be used a simple mechanism to form an approximate
inverse of a preconditioner. To build the factorization, the sparse
upper and lower triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are
computed such that the residual matrix formed by the factorization
\cite{saad_iterative_2003}:
\begin{equation}
  \mathbf{R} = \mathbf{L} \mathbf{U} - \mathbf{A} \:,
  \label{eq:ilu_residual_matrix}
\end{equation}
has a specified sparsity pattern and element magnitude. When a
magnitude threshold is used, elements generated in the factors below
that magnitude are dropped, resulting in ILU threshold (ILUT)
preconditioning. The sparsity pattern in this case is determined from
the input matrix to be preconditioned and the number of elements
maintained in the factor is specified by a fill level parameter. A
fill level of 1 will generate the same number of elements as the
sparsity pattern of the input matrix while a fill level of 2 will
contain twice as many elements in the factorization, resulting in a
better representation of the true LU factorization. The inverse of the
lower and upper triangular factors may then be easily inverted by
means of simple elimination to produce the preconditioner. For MCSA,
$\mathbf{L}^{-1}$ will be used on the left and $\mathbf{U}^{-1}$ on
the right to precondition the system.

@@

\subsection{Sparse Approximate Inverse Preconditioning}
\label{subsec:spn_spainv_preconditioning}
SPAINV is chosen next because it the true inverse of the
preconditioner is product of the algorithm, avoiding the need for many
inverse matrix applications. In addition, the algorithm seeks to
maintain the sparsity of the inverse preconditioner, potentially
alleviating some of the dense operator issues observed with
ILUT. Explicit formation of the preconditioner inverse is a
requirement of the current MCSA preconditioning strategy. As a result,
although ILUT preconditioning provided excellent iterative
performance, a significant time penalty is paid to extract the inverse
of the triangular factors through $N$ triangular solves. In addition,
this explicit inversion yielded a composite linear operator that was
dense, requiring modification of the system in order to achieve better
scalability and Monte Carlo performance.

Sparse approximate inverse (SPAINV) preconditioning is a technique
that may potentially alleviate both of these constraints at the cost
of reduced iterative performance. The method produces a sparse
approximation of the inverse matrix directly by minimizing the
Frobenius norm of the residual matrix \cite{saad_iterative_2003}:
\begin{equation}
  || \mathbf{I} - \mathbf{A} \mathbf{M} ||^2_F =
  \sum_{j=1}^N ||\mathbf{e}_j - \mathbf{A} \mathbf{m}_j||^2_2 \:,
  \label{eq:frobenius_norm_min}
\end{equation}
where the inverse preconditioning matrix $\mathbf{M}$ minimizes the
norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
$j^{th}$ columns of the identity and preconditioning matrices
respectively. Here, $\mathbf{M}$ is the actual inverse of the
preconditioner and is formed directly by the algorithm. On the right
hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
represented as a effective sum of $N$ linear system residual norms.

@@

\subsection{Applying the Reduced Domain Approximation}
\label{subsec:spn_prec_rda}
Although ILUT and SPAINV preconditioning permitted MCSA solutions for
the fuel assembly problem, a primary concern is the number of non-zero
states in each row of the system generated by the explicit
preconditioning strategy. In many cases, orders of magnitude more
matrix elements were generated resulting in poor scalability for
domain decomposed algorithms and overall performance issues for Monte
Carlo. As outlined in \S~\ref{subsec:reduced_domain_approximation},
the reduced domain approximation may be used as a mechanism to
potentially alleviate this problem by filtering elements of the
composite matrix in each row that fall below a certain threshold value
or by maintaining the largest $N$ elements in each row where $N$ is a
designated fill level. In this section, we will apply the reduced
domain approximation to both ILUT and SPAINV preconditioned system to
reduce the density of the composite linear operator to more manageable
levels and study the resulting iterative performance.

For each preconditioner, the parameters that achieved the best quality
metric results from the previous analysis were used. These correspond
to ILUT parameters of a fill level of 5.0 and a drop tolerance of
\sn{1}{-5} and SPAINV parameters of a sparsity level of 4 and a
threshold of 0.1. The reduced domain threshold was set to \sn{1}{-10}
in order to eliminate any exceedingly small values from the matrix
(often this is simply removing non-zero elements within the floating
point tolerance of zero). The reduced domain fill level was then
varied, starting with the largest non-zero entries per row value
observed for each of the preconditioning types in order to assess its
effects relative to the case where no reduced domain approximation was
applied.

Figure~\ref{fig:rda_iterations} gives the number of iterations
required to converge for each preconditioner type as a function of
reduced domain fill level. Figure~\ref{fig:rda_quality} gives the
corresponding quality metric for each data point where the number of
non-zero entries used to compute the metric is equivalent to the
reduced domain fill level. We first note that SPAINV preconditioning
alone is significantly more sensitive to the reduction in domain size
over the ILUT-based methods, although the preconditioner was of
$O(100)$ non-zero entries per row without any approximation
applied. For the ILUT-based methods, performance was significantly
better with convergence achieved in less than 40 MCSA iterations with
only 10 non-zero entries in each row (vs. 7 non-zero entries for the
case with no preconditioning). Looking at the quality metric data, we
see a reduction in the quality metric as a function of the reduced
domain fill level, achieving 2 orders of magnitude reduction in the
quality metric for the ILUT preconditioning.

@@

Figure~\ref{fig:relax_iters} gives the number of iterations required
to converge the fuel assembly problem for a 1-group SP1 discretization
using varying combinations of the relaxation parameters. First, both
parameters were fixed at the base case of 1 while the other parameter
was varied as shown in the left-hand side plots of
Figure~\ref{fig:relax_iters}. For the Richardson relaxation parameter,
using a value larger than 1 gave better iterative performance up to a
point, effectively providing a stronger extrapolation using the
residual at each iteration. For the Neumann relaxation parameter, a
value of less than 1 is observed to be ideal. Although initially
counter-intuitive, the fact that the correction computed by the Monte
Carlo solver has a stochastic error associated with it means that by
using a Neumann relaxation parameter less than 1, the correction and
its error are effectively dampened to improve iterative
performance.

Considering the CPU time required to converge a single eigenvalue
iteration, similar results are also observed for the relaxation
parameters as given by Figure~\ref{fig:relax_time}. For the base cases
given by the plots on the left, it was found the a Richardson
relaxation parameter of 1.1 and a Neumann relaxation parameter of 0.7
provided the fastest CPU time for convergence. Fixing each parameter
at these new values, the calculations were repeated as shown for the
plots on the right hand sides of both Figures~\ref{fig:relax_iters}
and \ref{fig:relax_time}. For each repeated timing calculation, it was
found that the same combination of relaxation parameters found in the
base cases performed the best although they did not necessarily have
the best iterative performance.

@@

\subsection{MCSA Estimator and Fixed Point Iteration Analysis}
\label{subsec:spn_estimator_comparison}
The iterative performance of MCSA will next be studied with both the
collision and expected value estimators using the best preconditioning
and relaxation parameter combinations found by the previous
analysis. Additionally, both the Richardson-based MCSA iteration given
by Eq~(\ref{eq:mcsa}) and the minimal residual-based MCSA iteration
given by Eq~(\ref{eq:mcsa_min_res}) will be used. Although the
expected value estimator achieved the best iterative performance in
the previous comparison and analysis for the simple Poisson problem in
\S~\ref{sec:parameter_estimator_analysis}, it is possible that the
collision estimator may perform better from a timing perspective. The
deterministic component of the expected value estimator that couples
the current state to other local states through the iteration matrix
stencil requires application of the estimator to potentially several
orders of magnitude more states at every transition event.

@@

Figure~\ref{fig:spn_estimator_iters} gives the iteration count results
and Figure~\ref{fig:spn_estimator_time} gives the timing
results. Remarkably, although the fuel assembly problem is
significantly more complicated than the simple Poisson problem studied
in \S~\ref{sec:parameter_estimator_analysis}, the results here are
largely the same with Figure~\ref{fig:spn_estimator_iters} and
Figure~\ref{fig:estimator_nh_iters} having identical qualitative
behavior. Using the expected value estimator with both fixed point
iterations again creates a relative insensitivity to the number of
histories used to compute the correction. Marginally better iterative
and timing performance was observed for the expected value estimator
when using the Richardson iteration. Better timing performance is
expected as computing the extrapolation parameter in the minimal
residual iteration requires several additional parallel
operations.

Compared to the collision estimator, the minimal residual iteration
permits fewer stochastic histories to be used with each MCSA iteration
while still maintaining good iterative performance. With respect to
CPU time, there is a more pronounced effect due to the larger density
of states created by the explicit ILUT preconditioning. Using a
reduced domain fill level of 100 creates a situation where the
expected value estimator is significantly more expensive per history
to compute that the collision estimator due to the coupling of states.
For the collision estimator, using around 20,000 stochastic histories
at every iteration gave the best iterative performance while 500
histories per iteration gave the best iterative performance when using
the expected value estimator. For purely serial computational
performance, there is not a distinguishing feature for the fuel
assembly problem presented that would cause one to choose one of the
estimators and fixed point iterations over the other.

@@

@@

%%---------------------------------------------------------------------------%%
\section{MCSA Verification\ }
\label{sec:spn_mcsa_verification}
Through the numerical studies presented in this chapter, we have
developed a set of preconditioning techniques that permit MCSA to be
used with the $SP_N$ form of the neutron transport equation and
applied to a difficult fuel assembly criticality problem. Along with
these preconditioning techniques, a set a varying MCSA parameters were
analyzed to find those that yielded the best performance for this
particular problem. In addition, several steps were taken to mitigate
the dense composite linear operators that arise when performing
explicit MCSA preconditioning necessary for achieve convergence. Using
these results, we now compare MCSA to conventional Krylov solvers that
would be typically used to solve the $SP_N$ form of the transport
equation as presented here.

@@

Table~\ref{tab:serial_differences_g1},
Table~\ref{tab:serial_differences_g2}, and
Table~\ref{tab:serial_differences_g4} give the infinity norms and
2-norms of the flux relative difference vectors given by
Eq~(\ref{eq:relative_difference_vector}) for all calculations and all
energy groups. In general, we see good agreement in the 2-norm between
the MCSA variants and the GMRES calculation, meaning that they have
approximately the same variation from the BiCGStab calculation. We do
note larger infinity norm values of this vector for some of the MCSA
variants relative to the GMRES results (i.e. group 1 in the 2-group
calculation), however, this large maximum value is apparently a local
phenomenon with good agreement in the 2-norm for the
calculations. Based on the good agreement between MCSA and GMRES for
the relative difference of the group fluxes and the eigenvalue results
presented here, the MCSA implementation should be deemed correct for
the fuel assembly problem for serial calculations with multiple energy
groups.

@@

%%---------------------------------------------------------------------------%%
\section{MCSA Performance Comparison to Conventional Methods\ }
\label{sec:spn_comparison}
With the same set of solver parameters used for verification
caluclation, MCSA will now be compared to the same Krylov solvers in
terms of both iterative performance and CPU timing. A Richardson
iteration solution will also be used in order to assess the
acceleration provided by the residual Monte Carlo component of MCSA.
To compare iterative performance, the number of linear solver
iterations required to converge each eigenvalue iteration was recorded
and then averaged over all eigenvalue iterations for each
solver.

Table~\ref{tab:spn_comparison_iterations} gives the average number of
linear solver iterations required to converge a single eigenvalue
iteration of the fuel assembly problem as a function of the number of
energy groups. In general, the iterative performance of all the
solvers tested aside from the Richardson iteration is comparable with
BiCGStab performing the best. We expect this because not only are the
multigroup $SP_N$ equations positive-definite, but they are also
block-symmetric and therefore we expect the conjugate gradient-based
methods to perform well due to the structure of the projection
scheme. It important to note that the iterative performance of each
solver is not a strong function of the number of energy groups in the
problem (and therefore the problem size). Even more important,
although BiCGStab produced the best iterative results, MCSA did
perform better than GMRES in many cases with the Richardson iteration
clearly accelerated by the Monte Carlo solve. When compared to the
preconditioned Richardson iteration, the acceleration provided by MCSA
is clear with 3-4 times fewer iterations required for convergence.

@@

\begin{itemize}
\item MCSA can solve the asymmetric system generated by the $SP_N$
  equations
\item MCSA has been incorporated into the Exnihilo neutronics
  production code base developed at Oak Ridge National Laboratory
\item Simple Jacobi-based preconditioning reduces the spectral radius
  of the $SP_N$ system below unity, however, this alone is not
  sufficient for MCSA convergence
\item As the spectral radius approaches unity, MCSA was demonstrated
  to break down with more stochastic histories required for
  convergence and more CPU time required per history
\item Light water reactor problems are difficult to solve with MCSA as
  they have large spectral radii due to the neutron scattering in the
  moderator
\item Advanced algebraic preconditioning strategies were applied to
  the $SP_N$ equations to obtain convergence with ILUT chosen for
  subsequent investigations
\item The reduced domain approximation was required to mitigate the
  memory and performance constraints of the explicit preconditioning
  strategy although both remain significant obstacles
\item MCSA relaxation parameters were developed and observed to
  enhance the time to solution for the fuel assembly criticality
  problem
\item MCSA was verified to produce the same k-eigenvalue for the fuel
  assembly as production Krylov methods and gave good relative flux
  agreement
\item MCSA was observed to converge in fewer iterations per eigenvalue
  iteration than GMRES for the fuel assembly criticality problem and
  more than Bi-CGStab using the same preconditioning
\item The explicit preconditioning strategy required to overcome the
  MCSA spectral radius restriction creates runtimes $O(100)$ larger
  than the production Krylov methods for the fuel assembly criticality
  problem
\end{itemize}
